---
title: "Intro to Data Frames in R"
subtitle: "Environmental Data Analytics"
author: "Luana Lima"
date: "Fall 2023"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length.

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

```{r cars}
cars
```

Note that it has 2 columns and 50 rows.

## Data frame columns

Suppose you want just the column speed. How would you access that data?

```{r}
cars$speed #syntax to pull one column is <chartname>$<columnname>
mtcars$cyl
```

How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```

## Transforming object in a data frame

Is the new object you create a data frame?

```{r}
#Option 1
class(car_speed)

#Option 2
is.vector(car_speed)
```

How could you make it a data frame?

```{r}
df_car_speed <- as.data.frame(car_speed) #taking vector and treating it as data frame
df_car_speed
class(df_car_speed)
```

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

```{r}
car_dist <- cars$dist #creating new variable

#Option 1
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind(), binding columns creates a dataframe
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist)
class(df_opt2)
df_opt2

```

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names() they do the same thing
names(df)
```

## Creating a data frame

How would you create a data frame?

```{r}
#useful function data.frame()
create_df <- data.frame("speed"=car_speed,"dist"=car_dist)
create_df
```

## Data frame functions

Some useful functions to use with data frames.

```{r}
ncol(df)  
nrow(df)  
length(df)  #same as ncol
summary(df) #creates summary stats of the dataset, helpful for exploring data quickly
head(df,6) #show the first 6 rows of df

#If you know the number of the column you want you can refer to that to access column
df[,1] #this is called Matrix Notation, specify the [row,col], if no row entered it will give you ALL of them
df[1,] #this will give you all columns for row 1
df[1:10,2] #get rows 1 thorugh 10 in column 2
df[c(1,2,3),] #create a matrix, c() calls a VECTOR to get multiple rows
df[c(2,1,3),]$car_speed #if you don't know what NUMBER of col, you can use the $<name> to call the correct column
df[1,2]

#you could also use this notation to delete columns
df <- df[,-2]
df
```
