---
output: html_document
editor_options: 
  chunk_output_type: console
---
## Objectives
1. Import and explore datasets in R
2. Graphically explore datasets in R
3. Apply data exploration skills to a real-world example dataset

## Opening discussion: why do we explore our data?

Why is data exploration our first step in analyzing a dataset? What information do we gain? How does data exploration aid in our decision-making for data analysis steps further down the pipeline?

## Import data and view summaries

```{r, message = FALSE}
# 1. Set up your working directory
getwd()

# 2. Load packages
library(tidyverse)

# 3. Import datasets
USGS.flow.data <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv",stringsAsFactors = T)

#View(USGS.flow.data)
# Alternate option: click on data frame in Environment tab

colnames(USGS.flow.data)
str(USGS.flow.data)
dim(USGS.flow.data)

# Check our date column
class(USGS.flow.data$datetime)

USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, format = "%Y-%m-%d") 
class(USGS.flow.data$datetime)

```


## Visualization for Data Exploration

Although the `summary()` function is helpful in getting an idea of the spread of values in a numeric dataset, it can be useful to create visual representations of the data to help form hypotheses and direct downstream data analysis. Below is a summary of the useful types of graphs for data exploration.

Note: each of these approaches utilize the package "ggplot2". We will be covering the syntax of ggplot in a later lesson, but for now you should familiarize yourself with the functionality of what each command is doing.

### Bar Chart (function: geom_bar)

Visualize count data for categorical variables. 

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data, aes(x = discharge.mean.approval)) +
  geom_bar()
 #aes is the aesthetics of HOW we want the plot to look
#this plot makes it easier for us to visualize categorical data to get an understanding of what we are dealing with

ggplot(na.omit(USGS.flow.data), aes(x = discharge.mean.approval)) +
  geom_bar()
#results in way less data

```

### Histogram (function: geom_histogram)

Visualize distributions of values for continuous numerical variables. What is happening in each line of code? Insert a comment above each line. Use with NUMERIC columns

```{r, fig.height = 3, fig.width = 4}
#Specifying a creation of a histogram showing the numeric vals
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean))

#Same BUT appl;ying additional arg of 'binwidth' to go beyond default and go to 10; makes it smoother
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), binwidth = 10)

#this chaging the bins again, makes it less smooth. Number of bins usually goes bw 5 and 20, but you need to figure out what PATTERN you are trying to show
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), bins = 20)

#continue to show discharge mean and highlight a SUBSET of the means
ggplot(USGS.flow.data, aes(x = discharge.mean)) +
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(limits = c(0, 500)) #this command  changes the x-axis to help "zoom in" to your histogram
  
#showing a diff value column that shows a peak that is PAST zero on gauge height mean and shows soemthing a little different than the discharge mean data did
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean), bins = 40)

```
### Frequency line graph (function: geom_freqpoly)

An alternate to a histogram is a frequency polygon graph (distributions of values for continuous numerical variables). Instead of displaying bars,  counts of continuous variables are displayed as lines. This is advantageous if you want to display multiple variables or categories of variables at once.

```{r, fig.height = 3, fig.width = 4}
#showing the dist of freq of different parts of our data
ggplot(USGS.flow.data) + #supplying data
  #geom_histogram(aes(x = gage.height.mean), bins = 50) #makes a histogram
  geom_freqpoly(aes(x = gage.height.mean), bins = 50) + #makes a line 
  geom_freqpoly(aes(x = gage.height.min), bins = 50, color = "red") + #changes color of specific line vs in aes would be changing by category
  geom_freqpoly(aes(x = gage.height.max), bins = 50,  lty = 2) + #lty is line type
  scale_x_continuous(limits = c(0, 10))

#adding another aes category: setting a COLOR to be different based on a CATEGORY of data 
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean, color = gage.height.mean.approval), bins = 50) + #got 'A' and 'P' by reading into gage.height.mean.approval
  scale_x_continuous(limits = c(0, 10)) +
  theme(legend.position = "top")

```
### Box-and-whisker plots (function: geom_boxplot, geom_violin)

A box-and-whisker plot is yet another alternative to histograms (distributions of values for continuous numerical variables). These plots consist of: 

* A box from the 25th to the 75th percentile of the data, called the interquartile range (IQR).

* A bold line inside the box representing the median value of the data. Whether the median is in the center or off to one side of the IQR will give you an idea about the skewness of your data.

* A line outside of the box representing values falling within 1.5 times the IQR. 

* Points representing outliers, values that fall outside 1.5 times the IQR. 

An alternate option is a violin plot, which displays density distributions, somewhat like a hybrid of the box-and-whiskers and the frequency polygon plot.

```{r, fig.height = 3, fig.width = 4}
#'x' is categorical value, 'y' is parsing into 2 data sets and showing summary stats on two things; box shows 1.5 times IQR and contains the BULK of the data, all others points are outliers
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean.approval, y = gage.height.mean))

#taking diff slices of data and looking at the dist which is done with the 'group' command. sorta like a histogram but instead of counts we are looking at distribution of values within the bins
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean, y = discharge.mean, group = cut_width(gage.height.mean, 1)))

#violin plots shows range AND distribution within that range. compare to first box plot we made. Width of violin is HOW MANY data points are in that area
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75))
```

### Scatterplot (function: geom_point)
Visualize relationships between continuous numerical variables.

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data) +
  geom_point(aes(x = discharge.mean, y = gage.height.mean)) #shows that as we get more volume, the gauge height doesn't inrease as fast and this is because the sides of stream rise faster than the middle of the stream. Each POINT is a ROW in our data table

ggplot(USGS.flow.data) +
  geom_point(aes(x = datetime, y = discharge.mean)) #this one shows mean discharge at different times. also shows the break where we do NOT have data
```

Question: under what circumstances would it be beneficial to use each of these graph types (bar plot, histogram, frequency polygon, box-and whisker, violin, scatterplot)?

> Answer: 

## Ending discussion

What did you learn about the USGS discharge dataset today? What separate insights did the different graph types offer?
> Answer: 


How can multiple options for data exploration inform our understanding of our data?

> Answer: 


Do you see any patterns in the USGS data for the Eno River? What might be responsible for those patterns and/or relationships?

> Answer: 


