---
title: "Lab 3 - Data Types | Exploring Datasets"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2023"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

1.  Discuss and navigate different data types in R
2.  Create, manipulate, and explore datasets
3.  Date objects

## Data Types in R

R treats objects differently based on their characteristics. For more information, please see: <https://www.statmethods.net/input/datatypes.html>.

-   **Vectors** 1 dimensional structure that contains elements of the same type.
  (all numbers, or characters, or logical)

-   **Matrices** 2 dimensional structure that contains elements of the same type.
  (simialr to DF but do NOT accept columns of diff data types)

-   **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

-   **Lists** Ordered collection of elements that can have different modes. You can have a matrix, an array, a DF, etc.

-   **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (\# at beginning of line, ---- at end)

```{r Create various data structures}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] # to access an ELEMENT in a vector you put the number of WHERE you want to pull. aka 3 is the going to pull the third element

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # provide the matrix with the number of rows:num of cols, and what number of rows/cols you want it to make
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "quatro", "cinco"), 
                                  c("un", "deux", "trois", "quatre"))) #

matrix1[4, ] # wants 4th row, all columns
matrix1[ , 3] # wants all rows and the third col
matrix1[c(12, 14)] # this calls index 12 and 14 of the matrix, without the c may be more useful
matrix1[c(12:14)] # this calls all elements between the 12th and 14th element
matrix1[2:4, 1:3] # this creates a new matrix with those specific ranges

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # 
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]]

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? NUMERIC
e <- c("red", "white", "red", NA) # What type of vector? CHARACTER
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? LOGICAL
dataframe1 <- data.frame(d,e,f) # creates a dataframe that is combining d, e, and f vectors
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # renames the col name for the three variables

dataframe1[1:2,] # this calls the first 2 rows and all the cols
dataframe1[c("ID","Passed")] # 
dataframe1$ID #dollar sign means you are accessing that specific part of the df

```

Question: How do the different types of data appear in the Environment tab?

> Answer:

Question: In the R chunk below, write "dataframe1\$". Press `tab` after you type the dollar sign. What happens?

> Answer:

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r DataExercise-1a}
high_temp_f <- c(81, 83, 83, 84, 78, 71, 79, 82, 83, 84)
low_temp_f <- c(55, 56, 58, 60, 61, 63, 63, 64, 58, 60)
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius. Use a function to create the two new vectors from your existing ones in Fahrenheit.

```{r DataExercise-1b}
f_to_c_fctn <- function(temp_f){
  (temp_f - 32) * (5/9)
}

high_temp_c <- f_to_c_fctn(high_temp_f)
low_temp_c <- f_to_c_fctn(low_temp_f)

high_temp_c #these two things make sure it prints in the code
low_temp_c
```

Combine your four vectors into a data frame and add informative column names.

```{r DataExercise-1c}
df_temp <- data.frame(high_temp_f, high_temp_c, low_temp_f, low_temp_c)

```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r DataExercise-1d}
summary (df_temp)
sd (df_temp) #this will NOT work bc you can't apply sd to a data frame, you must ask it to do it for a certain column
sd(df_temp$high_temp_f) #use the $ to call out the specific one
```

### Date objects

Remember formatting of dates in R:

%d day as number (0-31) %m month (00-12, can be e.g., 01 or 1) %y 2-digit year %Y 4-digit year %a abbreviated weekday %A unabbreviated weekday %b abbreviated month %B unabbreviated month

```{r Coding with dates}
# Adjust date formatting for today
# Write code for three different date formats. 
# An example is provided to get you started.
# (code must be uncommented)
today <- Sys.Date()
format(today, format = "%B") # prints 'September'
format(today, format = "%m") # prints '09'
format(today, format = "%d-%m") # prints '18-09'
format(today, format = "%A-%b-%Y") # prints 'Monday-Sep-2023'

```

### Package lubridate

Install and load the package lubridate into your R session. Lubridate offers fast and user friendly parsing of date-time data. Create a string for today's data and then convert it to R date object using lubridate.

More info on lubridate [here][<https://cran.r-project.org/web/packages/lubridate/lubridate.pdf>].

```{r DateExercise1}
#install.packages("lubridate")
library(lubridate)

#Ex1
str_today <- "2023-feb-7" #this shows up in the env with QUOTES which means it is character
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today) #this shows up in env in the date format
date_obj_today

#Ex2
str_today2 <- "Feb 7, 2023"
#Since the format is month-day-year we will use function mdy()
date_obj_today <- mdy(str_today2)
date_obj_today

#Ex_3 - on your own...
str_juneteenth <- "19 June 1865"
#Since the format is month-day-year we will use function dmy()
date_juneteenth <- dmy(str_juneteenth)
date_juneteenth
```

```{r DateExercise2}
#century issue
str_past <- "55-feb-3" #two digit year assumes we are in 2000s, not the 1900s or before
date_obj_past <- ymd(str_past)
date_obj_past

#Build a function to fix year that is more general than the one discussed in the lesson
fix.early.dates <- function(d, cutoff) {
       m <- year(d) %% 100  #operator %% is a modular division i.e. integer-divide year(d) by 100 and returns the remainder 
       year(d) <- ifelse(m > cutoff, 1900+m, 2000+m)  #this will update year(d), year() is a function that returns the year for a data object
       return(d)
}

fixed_date_obj_past <- fix.early.dates(date_obj_past,cutoff=23) #cutoff could be the current year to be more general or any other depending on data set 
fixed_date_obj_past
```

```{r centuryfix}
#Fix for century issue
str_past <- "55-feb-3"
#Alternative 1
date_obj_past <- fast_strptime(str_past,"%y-%b-%d",cutoff_2000=23L) #cutoff_2000 is an argument name
date_obj_past

#Alternative 2
date_obj_past2 <- parse_date_time2(str_past,"ymd",cutoff_2000=23L)
date_obj_past2

#Functions ymd(), mdy(), ydm() do not take argument cutoff_2000
```

In some cases when dates are provided as integers, you may need to provide an origin for your dates. For example, excel date could be given as number of days since an origin date. Origin date can be different. When R looks at dates as integers, its origin is January 1, 1970. Check if that is true on your machine.

```{r dateorigin}
#Check if "1970-01-01" is your origin date.
lubridate::origin

```
