---
title: "Lab 4: Factors"
author: "Environmental Data Analytics | John Fay and Luana Lima "
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---


```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
#Read packages
library(tidyverse);library(lubridate); library(here)

#Note: `here` package does't use autocomplete, best to type in the cosole and then copy-paste into the rmd file

#Read in data, don't convert strings to factors
sites <- read.csv(here('Data/Raw/NWIS_SiteFlowData_NE_RAW.csv')) # you can separate by commas instead too; did NOT include 'stringAsFactors' argument
dim(sites)

#Convert date column to date object
sites$date <- ymd_hms(sites$date) #year, month, day, hours, minutes, seconds using the LUBRIDATE function

```


## Examine the lakename column as a character column
* Tip - Lubridate cheatsheet: <https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf>
```{r create and explore days of week column}

#Compute the day of the week each sample was collected
sites$dow <- wday(sites$date)
class(sites$date)

## What class is the dow column?
class(sites$dow)

#List the unique values in the column
unique(sites$dow)

#Summary - is it meaningful?
summary(sites$dow)
#not all numbers are actual real values, they could also be representative of names (or days of the week as seen here)
```

## Convert the days of the week to a factor
```{r convert dow column to factor and explore}

#Converting a vector to a factor
sites$dow <- factor(sites$dow)

## What class is the dow column now? --> factor with 3 levels
class(sites$dow)

#List the unique values in the column --> Levels show all the categories, even if there are no data points attached to it
unique(sites$dow)

#Summary - is it meaningful? --> gives list of unique values and the count for each
summary(sites$dow)

#Show the levels associated with our factor column
levels(sites$dow)
```

## Factors & Levels
A **factor** is a grouping variable: rows with the same value in the factor column are members of the same group. For example, all rows with a `dow` of 3 were collected on a Tuesday.

**Levels** are the unique values that the factors can take. The `unique()` function reveals the levels as does the `levels()` function. By default, the levels includes only the values in the factor column. However, we can add additional levels and we can also re-order levels. Why would we want to do this? Well, one example would be if we wanted to plot the number of samples recorded each day of the week, and in that plot we wanted to include all days of the week, not just the one's found in our data...

```{r plot samples recorded each day of the week}
#Create a bar plot showing the count of samples by day of week
ggplot(sites,aes(x=dow)) +
  geom_bar() #only shows tues wed and thurs which are the only levels that currently exist
```



```{r add levels to our data}
#Update the levels to include all seven days of the week (note we have to start with existing levels)
levels(sites$dow) <- c(levels(sites$dow),"1","2","6","7") #using levels command, call the original levels, creates a list that adds new levels to the list; this does NOT impact the data, only the LEVELS

#Confirm the change
levels(sites$dow)
summary(sites$dow)

#Plot again
ggplot(sites,aes(x=dow)) +
  geom_bar() + #first two lines make a chart that look the same as first one (only days that have vals)
  scale_x_discrete(drop=FALSE) #this adds the other levels, but still looks not good, so move to next block

```

We now have all days of the week, but the order is incorrect. Also, we have day numbers, not day names. How to fix?
```{r reorder and rename levels}

#Reorder by overwriting the column with the existing values assigned with a new order of levels
sites$dow <- factor(sites$dow, levels=c("1","2","3","4","5","6","7")) #Or... as.character(seq(1,7))

#See that the levels are updated
levels(sites$dow)

#View the plot
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE) #now the levels have been reordered (does NOT change data) and plot looks better

#Add **labels** to our factors to make numbers have the day of the week
#This is also an example of tidy code that he wants to see
sites$dow <- factor(
  sites$dow, 
  levels=c("1","2","3","4","5","6","7"),
  labels=c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")) #they need to have a 1:1 correspondance

#See that the levels are updated
levels(sites$dow)

#View the plot and now it makes sense and has all the days of the week!
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)
```

## An exercise for you: see if you can plot the number of samples recorded each month