---
title: '6 - Part 2: Interactive R Markdown'
author: Environmental Data Analytics | John Fay & Luana Lima 
date: "Fall 2023"
output: html_document
runtime: shiny
---

## Overview
RMarkdown is a powerful tool for creating dynamic documents that combine code, text, and graphics. With the addition of widgets, it's possible to make these documents even more interactive and engaging. In this lesson, we will explore how to create interactive HTML files using RMarkdown widgets.

## LESSON OBJECTIVES

1. Create plotting functions that accept variables
2. Use widgets from the `shiny` library to make interactive HTML documents

### Set up the coding environment
```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
#Load packages
library(tidyverse);library(here);library(lubridate)
install.packages("htmlwidgets")

#Load data
nutrient_data <- read.csv(
  here('Data/Processed_KEY','NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv'),
  stringsAsFactors = TRUE)

#Fix dates
nutrient_data$sampledate = ymd(nutrient_data$sampledate)

#Set a default plotting theme
my_theme <- theme_classic(base_size = 14)
theme_set(my_theme)

```


### Create a base plot
We will plot Total P as a function of time, coloring data by depth, and symbolizing by lake
```{r Create base plot, echo=FALSE, eval=FALSE}
#Construct the plot
nutrient_data %>% 
  filter(year4 > 1990 & year4 < 2005) %>% 
  ggplot(
    aes(
      x = sampledate, 
      y = tp_ug, 
      fill = depth_id, 
      shape = lakename)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_x_date(
    date_breaks = "1 years",
    date_labels = '%Y') +
  scale_shape_manual(values = c(21, 24)) +
  labs(
    title = "Concentration over time, by depth and lake",
    x = "Date", 
    y = expression(Concentration ~ (mu*g / L)), 
    shape = "Lake", 
    fill = "Depth ID") +
  scale_fill_viridis_c(
    option = "mako", 
    begin = 0, 
    end = 1, 
    guide = guide_colorbar(reverse = T),
    direction = -1)
```

## Using variables in plots
Now tweak the above code so that the variable we plot (`tp_ug`) is specified as a variable rather than hard coded in our ggplot command. We will also set the start and end dates to variables as well. 

To do this we need to convince R that the value we associate with the `y` aesthetic is a variable, not a column name. This is done by the odd convention of inserting the variable within the code `!!sym()`. The `!!` (pronounced "bang-bang") combined with `sym()` are "Tidy eval helpers" necessary because tidy-R does away with quotations. See more info [here](https://ggplot2.tidyverse.org/reference/tidyeval.html).

```{r Tweak plot to use variables, echo=FALSE}
#Creat variables
the_parameter = 'tp_ug'
start_year = 1990
end_year = 2005

#making a function for a plot makes it easier to only plot certain things 
plot_it <- function(the_parameter, start_year, end_year) {
    #Construct the plot

my_plot <- nutrient_data %>% 
  filter(year4 > start_year & year4 < end_year) %>% 
  ggplot(
    aes(
      x = sampledate, 
      y = !!sym(the_parameter), #need the !!sym to call the variable correctly
      fill = depth_id, 
      shape = lakename)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_x_date(
    date_breaks = "1 years",
    date_labels = '%Y') +
  scale_shape_manual(values = c(21, 24)) +
  labs(
    title = "Concentration over time, by depth and lake",
    x = "Date", 
    y = expression(Concentration ~ (mu*g / L)), 
    shape = "Lake", 
    fill = "Depth ID") +
  scale_fill_viridis_c(
    option = "mako", 
    begin = 0, 
    end = 1, 
    guide = guide_colorbar(reverse = T),
    direction = -1)

return(my_plot)
}

#exmaple of using the function to make a subplot

plot_it('tp_ug',1992,2005)


```

##Interactive R with Shiny Widgets
The R MarkDown Cheat Sheet provides a short example on how to make your Markdown document interactive. First you must alter your yaml section to read:  
```
output: html_document
runtime: shiny
```
Notice that when you do so, the Knit menu is replaced. 

Then we can add the code chunk listed in the document.Before running this code, however, we should set the code chunk options for the previous code chunks so that they are not run (eval) and not shown (echo). 

```{r eval=FALSE, include=FALSE}
numericInput(
  inputId = "n",
  label = "How many cars?",
  value = 5) #this makes variable start at 5

renderPrint(paste("You chose", input$n))

renderTable({
  head(cars, input$n) 
})
```

Shiny has many other input widgets: https://shiny.rstudio.com/gallery/widget-gallery.html. Below we'll use a Select Box widget to specifiy the variable shown in our plot. We'll also add a slider range to limit the date shown in our plot. 

Note: Before running this plot, set the previous code chunk to not evaluate

```{r widget exploration, echo=FALSE}
selectInput(
  inputId = "the_parameter", #enter the name of the variable we want
  label = "Select the parameter to plot", #how it shows up on the screen
  choices = c('tp_ug', 'tn_ug', 'no23', 'po4'),
  selected = 'tp_ug'
)

sliderInput(
  inputId = "YearSlider",
  label = "Select the date range",
  min = 1990,
  max = 2020,
  value = c(1995,2000) #the range it starts with on the slider
)

#renderPrint(input$YearSlider)


###This is not working for me.
renderPlot(
  plot_it(
    input$the_parameter,
)
  )

    # input$YearsSlider[1],
    # input$YearsSlider[2]
```


