---
title: "Assignment 2: Coding Basics"
author: "Maia Griffith"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq(1,30,3)  #first created a seq of numbers using seq (start, end, interval)
Thirty_by_3 <- seq(1,30,3) #assigning the seq to a descriptive name

#2. 
mean_seq <- mean(Thirty_by_3) #creating a variable for the mean and another for the median
med_seq <- median(Thirty_by_3)

#3. 
mean_seq > med_seq #then ask R to compare the two; TRUE (if mean is greater) or FALSE
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#a)
student_names <- c("Mary", "Carol", "Ryan", "Steve") #making a CHARACTER vector with 4 student names

#b) I used a random number generator to get 4 numbers and made a vector to contain them.
test_score <- c(88, 37, 49, 99) #NUMERIC vector

#c)
result<- (test_score > 50) #LOGICAL vector; returns TRUE/FALSE results

#7.
student_test_results <- data.frame(student_names, test_score, result) #created data frame

#8.changed column names in data frame 
colnames(student_test_results) <- c("Student_Name", "Test_Score", "Pass_or_Fail") 
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frames can have multiple classes of data while a matrix can only use one kind.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#10. creating my function
pass_or_fail_fctn <- function(test_score) {
  ifelse (test_score > 50, TRUE, FALSE) 
}

#11. testing my function
pass_or_fail_fctn(test_score)

#now trying 'if' and 'else' instead, function is able to be created BUT does not end up working
# if_else_fctn_test <- function(test_score) {
#   if (test_score > pass) {
#     print("True")
#   }
#     
#   else {
#     print("False")
#   }
#     
# }
# 
# if_else_fctn_test(test_score) 

#Result gets an error that says the condition has length > 1 (test_score is a vector of length 4)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The `ifelse` worked because it can be used for vectors, while if the `if` and `else` can only be used for a single variable and NOT for vectors.
