---
title: "Assignment 10: Data Scraping"
author: "Maia Griffith"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
# Load packages
library(tidyverse)
library(rvest)
library(lubridate)
library(here)

# Check directory
here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
# Using read_html from rvest package
LWSP_webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
WS_Name <- LWSP_webpage %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
PWSID <- LWSP_webpage %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
Ownership <- LWSP_webpage %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
MGD_values <- LWSP_webpage %>% html_nodes('th~ td+ td') %>% html_text()
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
# Converting into data frame and mutate in a Date column
water_df <- data.frame(
  "Water System" = WS_Name,
  "PWSID" = PWSID,
  "Ownership" = Ownership,
  "Monthly Max Day Use" = as.numeric(MGD_values),
  "Month" = c("Jan", "May", "Sept", "Feb", "June", "Oct", 
              "March", "July", "Nov", "April", "Aug", "Dec"),
  "Year" = rep(2022,12)
  ) %>%
  mutate(Date = my(paste(Month,Year)))

#5 
# Create a line plot across months for 2022
monthly_water_plot <- 
  ggplot(water_df,
         aes(x = Date,
             y = Monthly.Max.Day.Use)) +
  geom_line() + 
  labs(title = paste("2022 Max Daily Withdrawals per Month"),
       subtitle = "Durham, NC",
       y="Withdrawal (MGD)",
       x="Month") +
  scale_x_date(date_breaks = "1 month", date_labels = '%b')

monthly_water_plot
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape.PWSID <- function(the_PWSID, the_year){
  # Get the proper url/website
  the_url <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',the_PWSID,'&year=',the_year))

  #Set the element address variables (determined in the previous step)
  the_WS_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  the_ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  the_MGD_tag <- 'th~ td+ td'

  # Locate elements and read their text attributes into variables
  WS_Name <- the_url %>% html_nodes(the_WS_tag) %>% html_text()
  Ownership <- the_url %>% html_nodes(the_ownership_tag) %>% html_text()
  MGD_values <- the_url %>% html_nodes(the_MGD_tag) %>% html_text()

  # Construct a data frame from the values
  water_func_df <- data.frame(
    "Water System" = rep(WS_Name,12),
    "PWSID" = rep(the_PWSID,12),
    "Ownership" = rep(Ownership,12),
    "Monthly Max Day Use" = as.numeric(MGD_values),
    "Month" = c("Jan", "May", "Sept", "Feb", "June", "Oct",
            "March", "July", "Nov", "April", "Aug", "Dec"),
    "Year" = rep(the_year,12)
    ) %>%
    mutate(Date = my(paste(Month,Year)))

  # Return the dataframe
    return(water_func_df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
# Use the function for new data frame
df_Dur_2015 <- scrape.PWSID('03-32-010',2015)

# Create a line plot across months for Durham 2015
Dur2015_water_plot <- 
  ggplot(df_Dur_2015,
         aes(x = Date,
             y = Monthly.Max.Day.Use)) +
  geom_line() + 
  labs(title = paste("2015 Max Daily Withdrawals per Month"),
       subtitle = "Durham, NC",
       y="Withdrawal (MGD)",
       x="Month") +
  scale_x_date(date_breaks = "1 month", date_labels = '%b')

Dur2015_water_plot

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
# Get Ashville 2015 data
df_Ash_2015 <- scrape.PWSID('01-11-010',2015)

# Combine with Durham 2015 data
df_Dur_Ash_2015 <- inner_join(df_Dur_2015, df_Ash_2015)

DurAsh2015_water_plot <- 
  ggplot(df_Dur_Ash_2015) +
  geom_line(aes(x = Date,
             y = Monthly.Max.Day.Use,
             color = Water.System)) + 
  labs(title = paste("Ashville vs Durham Max Daily Withdrawals per Month"),
       subtitle = "2015 Data",
       y="Withdrawal (MGD)",
       x="Month") +
  scale_x_date(date_breaks = "1 month", date_labels = '%b')

DurAsh2015_water_plot  

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
#Create a list of the year we want, the same length as the vector above
the_years <- seq(2010,2020)

#"Map" the "scrape.it" function to retrieve data for all these
dfs_2020 <- map2(the_years, the_facility_ids_subset, scrape.it)

# Plot
ggplot(the_df,aes(x=Date,y=Avg_Withdrawals_mgd)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE)
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
>
