---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Maia Griffith"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 Setting up my session.
library(tidyverse)
library(agricolae)   
library(corrplot)
library(dplyr)
library(here)
here()

#Load the correct dataset
lake.chem.phys <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
                           stringsAsFactors = TRUE)
#Change into date format
lake.chem.phys$sampledate <- as.Date(lake.chem.phys$sampledate , format = "%m/%d/%Y")

#2 Build and set new theme.
MGtheme <- theme_light(base_size = 11) +
  theme(axis.text = element_text(color = "darkgray"), 
        legend.position = "right")
theme_set(MGtheme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: July mean lake temperature across all lakes does not change with depth. 
Ha: July mean lake temperature across all lakes changes with depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 Pipe to wrangle the data
Jul.lake.chem.phys <-
  lake.chem.phys %>% # Calling the data frame
  mutate(Month = month(sampledate)) %>% # Creating a month col
  filter(Month == 7 ) %>% # Filtering by July (month number 7)
  select(lakename, year4, daynum, depth, temperature_C) %>% # Selecting cols
  na.omit() # Removing ALL NA values from ALL cols

dim(Jul.lake.chem.phys)

#5 Scatter plot of temp by depth (x = depth, y = temp)
temp_depth_plot <-
  ggplot(Jul.lake.chem.phys, aes(x = depth, # Creating the plot
                                 y = temperature_C)) +
  geom_point() + # Scatterplot
  geom_smooth(method = "lm") + # Adding line of best fit
  labs(y = "Temperature (C)",
       x = "Depth (meters)",
       title = "Lake Temperature by Depth") + # Editing axis titles
  ylim(0,35) # Limiting temp values

temp_depth_plot

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot suggests that there is a negative relationship between temperature and depth, meaning with increasing depth there is decreasing temperature.The distribution of points suggests that the trend may not be simply linear, but have a curve of some kind in which temperature decreases strongly within the first 10 meters, and then starts to level out.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
tempdepth.regr <- 
  lm(Jul.lake.chem.phys$temperature_C ~ 
       Jul.lake.chem.phys$depth) 

summary(tempdepth.regr)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The Adjusted R-squared value is 0.7387, which means that about 73.8% of the variability in temperature is explained by changes in depth. There are 9726 degrees of freedom, with a high statistical significance with a p-value of less than 0.05 (p-value: < 2.2e-16). For every 1m change in depth, the temperature decreases by about 1.9 degrees C.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 AIC for year4, daynum, and depth for predicted temp
Jul.lakes.AIC <- lm(data = Jul.lake.chem.phys, 
                    temperature_C ~ year4 + daynum + depth)

step(Jul.lakes.AIC) 
# Stepwise AIC shows the initial AIC is best, with each step making it worse.
# It also shows that removing depth has the largest impact.
# The variables year4 has very little impact, and daynum has a small impact.

#10 New regression using all 3 variables to explain temperature.
temp.depthdaynumyear.regr <- lm(data = Jul.lake.chem.phys, 
                           temperature_C ~ year4 + depth + daynum)
summary(temp.depthdaynumyear.regr)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests we use year4 (low impact), depth (high impact) and daynum (some impact) in our multiple regression. This new model explains 74.1% of the variability in temperature, which is a small improvement over only using depth.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Format ANOVA as aov
Jul.Temp.anova <- aov(data = Jul.lake.chem.phys, temperature_C ~ lakename)
summary(Jul.Temp.anova)

# Format ANOVA as linear model (lm)
Jul.Temp.anova.lm <- lm(data = Jul.lake.chem.phys, temperature_C ~ lakename)
summary(Jul.Temp.anova.lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes, because the p-value (in both models) is less than 0.05 (p-value: < 2.2e-16).


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
temp_depth_ANOVA_plot <-
  ggplot(Jul.lake.chem.phys, aes(x = depth, 
                                 y = temperature_C,
                                 color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(y = "Temperature (C)",
       x = "Depth (meters)",
       color = "Lake Name",
       title = "Comparison of Temperature by Depth in Each Lake") +
  ylim(0,35)

temp_depth_ANOVA_plot

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Jul.Temp.anova)

#Then find groupings of lakes
Temp.groups <- HSD.test(Jul.Temp.anova, "lakename", group = TRUE)
Temp.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The HSD.test shows that Paul Lake and Ward Lake have a mean temperature that is statistically the same as Peter Lake. It does not look like any lake is statistically distinct from all other lakes.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two-sample t-test could be run since we want to test the hypothesis that the means of the two lakes are equivalent. A low p-vale of less than 0.05 would tell us that the means ARE statistically different from each other.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Wrangle the data
Cram.Ward.combo <- Jul.lake.chem.phys %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

# Used the following to make sure the filter worked properly
    #summary(Cram.Ward.combo$lakename)
    #summary(Jul.lake.chem.phys$lakename)

# Two-sample T-test (null: the lake means are equal; alt: means are not equal)
temp.twosample <- 
  t.test(Cram.Ward.combo$temperature_C ~ Cram.Ward.combo$lakename)
temp.twosample

```

>Answer: The results give a p-vale > 0.05, therefore we fail to reject the null hypothesis, and can conclude that there is not a signficant difference in the mean temperatures between the two lakes. This does match my number 16, HSD.test resulted in a grouping of Crampton and Ward lakes, which also meant that there was not a significant difference between the temperatures of those lakes. 
